A generalized quadratic estimate for random field nonstationarity \thanksref{code}

### 2016-03-02 paper/paper.tex
just a few typos

\begin{abstract}
In this paper, we attempt to shed light on a new class of nonstationary random fields which exhibit, what we call, {\em local invariant nonstationarity}. We argue that the local invariant property has a special interaction with a new  generalized quadratic estimate---also derived here---which extends an estimate originally developed for gravitational lensing of the Cosmic Microwave Background in Cosmology \cite{hu2001mapping, hu2002mass}. The nature of this interaction not only encourages low estimation bias but also enables accurate (and fast) quantification of Frequentist mean square error quantification of the estimated nonstationarity. These quadratic estimates are interesting, in their own right, as they detect and estimate nonstationarity by probing correlation among Fourier frequencies, the absence of which is the characterizing feature of weak stationarity (by Bochner's Theorem). Moreover, this generalized quadratic estimate can be computed with an Fourier characterization that runs in $\mathcal O(n\log n)$ time when observing the field on a uniform grid of size $n$ in $\Bbb R^d $. Finally, the work presented here partially addresses two other problems associated with the statistical theory of nonstationarity: 1) estimating the phase of a spatially varying modulated stationary random field and 2) identifying a larger class of nonstationary random fields which admit an extension of the quadratic estimator of gravitational lensing that extends the same attractive statistical properties.
\end{abstract}
% \begin{keyword}
% 	\kwd{}
% 	\kwd{}
% \end{keyword}
\end{frontmatter}

@@

There are two main contributions to the variability in $\hat\phi_{\bs\ell}$. The first source of variability is due to the   additive observational noise $N(\bs x)$. The second source, sometimes called \textit{shape noise} in Cosmology, is due to the baseline stationary fluctuations in $Z(\bs x)$ characterized by the autocovariance function $K(\bs x - \bs y, \bs 0)$.  The spectral density of this shape noise can be approximated by ${(2\pi)}^{d/2}C^{(0)}_{\bs \ell}$, which corresponds to the zero${}^\text{th}$ order approximation in (\ref{first order term, intro}), or by $C^{ZZm}_{\bs \ell}$ which denotes the  spectral density of $Z(\bs x)$ marginalized over the prior from Assumption \ref{Assumption 3}. Both these approximations can be used, within the formulas derived below, to give accurate approximation to the mean squared variability in $\hat\phi_{\bs \ell}$. However, in the cosmology literature on gravitational lensing, the marginal stationary model for $Z(\bs x)$ is typically used, rather than ${(2\pi)}^{d/2}C^{(0)}_{\bs \ell}$, for shape noise quantification.

@@

In this section we give an example of two stochastic processes with nearly the same values of $\bs C_{\bs k}^{(0)}, \bs C_{\bs k}^{(1)}$ and $\bs C_{\bs k}^{(2)}$, discussed above, but where one process is not locally invariant.  A quadratic estimate of nonstationarity is derived for both models and the resulting bias of each is compared. The main conclusion is that a small deviation from the locally invariant structure results in a bias that is orders of magnitudes larger than what is found in the local invariant model.

@@

% --------------------------------------------------------
%  claim
%---------------------------------------------------------
\begin{claim}[\textbf{Any pair of spectral densities with the same integral are attainable}]
\label{claim any two spec are attainable}
Let $d\geq 1$ be an integer, $t_0 > 0$ be a real number and $C_{\bs k}$, $\tilde C_{\bs k}$ be two spectral densities on $\Bbb R^d$ with finite second moments such that $\sigma^2 = \int_{\Bbb R^d}  C_{\bs k}d\bs k = \int_{\Bbb R^d} \tilde C_{\bs k}d\bs k$. Then there exists a vector field $\bs \eta_{\bs k}\colon \Bbb R^d \rightarrow \Bbb R^d$ which is $L_2(\Bbb R^d)$ integrable with respect to $C_{\bs k}d\bs k$ and  generates a one dimensional curve of spectral measures  $\{\lambda^{(t)}: t\in [0,t_0]\}$, defined by 
\begin{align}
\lambda^{(t)} := (\bs k + t\bs \eta_{\bs k})\,\sharp\, C_{\bs k}
\end{align}
with endpoints $\lambda^{(0)}=C_{\bs k}d\bs k$ and $\lambda^{(t_0)}=\tilde C_{\bs k}d\bs k$,
such that  $\{\lambda^{(t)}:t\in[0,t_0] \}\subset \mathscr C^{C,\bs\eta}$. In particular there exists a nonstationary phase random field model for which  $C_{\bs k}$ and $\tilde C_{\bs k}$ are both locally attainable.
Moreover,  $\{\lambda^{(t)}: t\in [0,t_0]\}$ is a $L_2$-Wasserstein geodesic path within the class of absolutely continuous spectral measures (with total mass $\sigma^2$ and finite second moments) and each measure $\lambda^{(t)}$ has a density $C^{(t)}_{\bs k}$ with respect to Lebesgue measure on $\Bbb R^d$ which (weakly) satisfies 
\begin{align}
\label{div flow for geo}
\partial_t C^{(t)}_{\bs k} +\text{\rm div}(\bs \eta^{(t)}_{\bs k}C^{(t)}_{\bs k}) = 0
\end{align}
for all $t\in [0,t_0]$ where $\bs \eta^{(t)}_{\bs k}:=\bs \eta_{T_t^{-1}(\bs k)}$ and $T_t(\bs k):= \bs k + t\bs \eta_{\bs k}$.
\end{claim}
% --------------------------------------------------------
%  proof
%---------------------------------------------------------
\begin{proof}
By standard optimal transport theory (see \cite{villani2003topics} for example) the assumptions on $C_{\bs k}$ and $\tilde C_{\bs k}$ guarantee the existence of a convex function $\psi_{\bs k}:\Bbb R^d \rightarrow \Bbb R$ such that $\nabla \psi_{\bs k}$ is the optimal transport from $C_{\bs k}d\bs k$ to $\tilde C_{\bs k}d\bs k$.  Let $\bs \eta_{\bs k}:= \frac{1}{t_0}(\nabla \psi_{\bs k}-\bs k)$ so that
\begin{align}
\label{geo eq number 1}
\lambda^{(t)}:=(\bs k + t\bs\eta_{\bs k})\,\sharp\, C_{\bs k} = \big((1-\textstyle\frac{t}{t_0})\bs k + \textstyle\frac{t}{t_0}\nabla\psi_{\bs k}\big)\,\sharp\, C_{\bs k}. 
\end{align}
The particular form of the right hand side of (\ref{geo eq number 1})  implies each measure $\lambda^{(t)}$ has a density $C_{\bs k}^{(t)}$ with respect to Lebesgue measure and the path of measures $\{\lambda^{(t)}\colon t\in[0,t_0] \}$ forms an $L_2$-Wasserstein geodesic with endpoints $C_{\bs k}d\bs k$ and $\tilde C_{\bs k}d\bs k$ at $t=0$ and $t=t_0$ respectively (see Proposition 5.9 in \cite{villani2003topics}). Moreover, $C^{(t)}_{\bs k}$ weakly satisfies (\ref{div flow for geo}) by Theorem 5.34 of \cite{villani2003topics} and the fact that $C^{(t)}_{\bs k} = T_t\,\sharp\, C^{(0)}_{\bs k}$ where $\partial_t T_t(\bs k) = \bs \eta_{\bs k}$.
By  setting $\bs A=t\bs I_d$ in (\ref{eq: local spectral density for nonstationary phase}) on has
\[
\int  \exp\!\big(i(\bs x-\bs y)\!\cdot\! (\bs k + t\bs\eta_{\bs k})\big)\, C_{\bs k}\frac{d\bs k}{(2\pi)^d} 
=\int \exp\!\big(i(\bs x-\bs y)\!\cdot\! \bs\omega\big)\,  C^{(t)}_{\bs \omega}\frac{d\bs \omega}{(2\pi)^d}
\]
which implies that for each $t\in[0,t_0]$ the measure $\lambda^{(t)}$ is a locally attainable spectral measure.
\end{proof}

@@

% --------------------------------------------------------
%  proof
%---------------------------------------------------------
\begin{proof}
By the proof of Claim \ref{claim any two spec are attainable} it will be sufficient to show that $\tilde F^{-1}\circ F(|\bs k|)\frac{\bs k}{|\bs k|}$ is the optimal transport from $C_{|\bs k|}$ to $\tilde C_{|\bs k|}$.
Let $\bs K$ and $\tilde {\bs K}$ be random vectors in $\Bbb R^d$ with densities $C_{|\bs k|}/\sigma^2$ and $\tilde C_{|\bs k|}/\sigma^2$, respectively. By the distributional rotational symmetry of $\bs K$ there exists a convex $\psi(r):\Bbb R^+ \rightarrow \Bbb R$ such that $\nabla (\psi(|\bs k|))=\psi^\prime(|\bs k|)\frac{\bs k}{|\bs k|}$ is the optimal transport from $\mathscr L\bs K$ to $\mathscr L\tilde{\bs K}$ (where $\mathscr L\bs K$ and $\mathscr L\tilde{\bs K}$ denotes the law, i.e. probability distribution, of $\bs K$). Also notice that $\psi^\prime(r)$ is the optimal transport from $\mathscr L |\bs K|$ to  $\mathscr L|\tilde{\bs K}|$ since $\psi(r)$ is convex and
\begin{align*}
P(\psi^\prime(|\bs K|)\leq r) = P\big(\big|\psi^\prime(|\bs K|)\textstyle\frac{\bs K}{|\bs K|}\big|\leq r\big) = P(|\tilde{\bs K}|\leq r).
\end{align*}
The optimal transport between two univariate random variables (see \cite{villani2003topics}) is given by the composition of the quantile function (of the target measure) and the cumulative distribution function (of the base measure). Therefore the optimal transport from $\mathscr L|\bs K|$ to $\mathscr L|\tilde{\bs K}|$ is given by $\tilde F^{-1}\circ F(r)$ where $F(r)=\frac{2\pi^{d/2}}{\sigma^2\Gamma(d/2)}\int_0^r \xi^{d-1}C_\xi d\xi$ and $\tilde F(r)=\frac{2\pi^{d/2}}{\sigma^2\Gamma(d/2)}\int_0^r \xi^{d-1}\tilde C_\xi d\xi$ are the cumulative distributions functions of $|\bs K|$ and $|\tilde{\bs K}|$, respectively. By the uniqueness of optimal transports one has $\psi^\prime(r)=\tilde F^{-1}\circ F(r)$ and therefore 
\[
\nabla (\psi(|\bs k|))=\psi^\prime(|\bs k|)\frac{\bs k}{|\bs k|} = \tilde F^{-1}\circ F(|\bs k|)\frac{\bs k}{|\bs k|}
\]
is the optimal transport from $\mathcal L\bs K$ to $\mathcal L\tilde{\bs K}$,
as was to be shown.
\end{proof}

@@

In this section give some informal guidance for specifying $\bs \xi_{\bs k}, C_{\bs k}$ and $\bs\eta_{\bs k}$ in the nonstationary spectral phase model (see Definition \ref{def: nonstate phase}). Since the quadratic estimate is  adept at detecting small departures from stationarity---partly due to the accuracy of the variance calculations and the speed at which the quadratic estimate can be simulated under a null stationary model---we focus on the situation where the statistician wants to estimate potential nonstationary extensions of a base stationary model. Further details of this approach can be found in the simulation sections \ref{section: nonstat example d=1} and \ref{section: nonstat example d=2}.

>>>>

\begin{abstract}
In this paper, we attempt to shed light on a new class of nonstationary random fields which exhibit, what we call, {\em local invariant nonstationarity}. We argue that the local invariant property has a special interaction with a new  generalized quadratic estimate---also derived here---which extends an estimate originally developed for gravitational lensing of the Cosmic Microwave Background in Cosmology \cite{hu2001mapping, hu2002mass}. The nature of this interaction not only encourages low estimation bias but also enables accurate (and fast) quantification of Frequentist mean square error quantification of the estimated nonstationarity. These quadratic estimates are interesting, in their own right, as they detect and estimate nonstationarity by probing correlation among Fourier frequencies, the absence of which is the characterizing feature of weak stationarity (by Bochner's Theorem). Moreover, this generalized quadratic estimate can be computed with a Fourier characterization that runs in $\mathcal O(n\log n)$ time when observing the field on a uniform grid of size $n$ in $\Bbb R^d $. Finally, the work presented here partially addresses two other problems associated with the statistical theory of nonstationarity: 1) estimating the phase of a spatially varying modulated stationary random field and 2) identifying a larger class of nonstationary random fields which admit an extension of the quadratic estimator of gravitational lensing that extends the same attractive statistical properties.
\end{abstract}
% \begin{keyword}
% 	\kwd{}
% 	\kwd{}
% \end{keyword}
\end{frontmatter}

@@

There are two main contributions to the variability in $\hat\phi_{\bs\ell}$. The first source of variability is due to the   additive observational noise $N(\bs x)$. The second source, sometimes called \textit{shape noise} in Cosmology, is due to the baseline stationary fluctuations in $Z(\bs x)$ characterized by the autocovariance function $K(\bs x - \bs y, \bs 0)$.  The spectral density of this shape noise can be approximated by ${(2\pi)}^{d/2}C^{(0)}_{\bs \ell}$, which corresponds to the zero${}^\text{th}$ order approximation in (\ref{first order term, intro}), or by $C^{ZZm}_{\bs \ell}$ which denotes the  spectral density of $Z(\bs x)$ marginalized over the prior from Assumption \ref{Assumption 3}. Both these approximations can be used, within the formulas derived below, to give accurate approximations to the mean squared variability in $\hat\phi_{\bs \ell}$. However, in the cosmology literature on gravitational lensing, the marginal stationary model for $Z(\bs x)$ is typically used, rather than ${(2\pi)}^{d/2}C^{(0)}_{\bs \ell}$, for shape noise quantification.

@@

In this section we give an example of two stochastic processes with nearly the same values of $\bs C_{\bs k}^{(0)}, \bs C_{\bs k}^{(1)}$ and $\bs C_{\bs k}^{(2)}$, discussed above, but where one process is not locally invariant.  A quadratic estimate of nonstationarity is derived for both models and the resulting bias of each is compared. The main conclusion is that a small deviation from the locally invariant structure results in a bias that is orders of magnitude larger than what is found in the local invariant model.

@@

% --------------------------------------------------------
%  claim
%---------------------------------------------------------
\begin{claim}[\textbf{Any pair of spectral densities with the same integral are attainable}]
\label{claim any two spec are attainable}
Let $d\geq 1$ be an integer, $t_0 > 0$ be a real number and $C_{\bs k}$, $\tilde C_{\bs k}$ be two spectral densities on $\Bbb R^d$ with finite second moments such that $\sigma^2 = \int_{\Bbb R^d}  C_{\bs k}d\bs k = \int_{\Bbb R^d} \tilde C_{\bs k}d\bs k$. Then there exists a vector field $\bs \eta_{\bs k}\colon \Bbb R^d \rightarrow \Bbb R^d$ which is $L_2(\Bbb R^d)$ integrable with respect to $C_{\bs k}d\bs k$ and  generates a one dimensional curve of spectral measures  $\{\lambda^{(t)}: t\in [0,t_0]\}$, defined by 
\begin{align}
\lambda^{(t)} := (\bs k + t\bs \eta_{\bs k})\,\sharp\, C_{\bs k}
\end{align}
with endpoints $\lambda^{(0)}=C_{\bs k}d\bs k$ and $\lambda^{(t_0)}=\tilde C_{\bs k}d\bs k$,
such that  $\{\lambda^{(t)}:t\in[0,t_0] \}\subset \mathscr C^{C,\bs\eta}$. In particular there exists a nonstationary phase random field model for which  $C_{\bs k}$ and $\tilde C_{\bs k}$ are both locally attainable.
Moreover,  $\{\lambda^{(t)}: t\in [0,t_0]\}$ is a $L_2$-Wasserstein geodesic path within the class of absolutely continuous spectral measures (with total mass $\sigma^2$ and finite second moments) and each measure $\lambda^{(t)}$ has a density $C^{(t)}_{\bs k}$ with respect to Lebesgue measure on $\Bbb R^d$ which (weakly) satisfies 
\begin{align}
\label{div flow for geo}
\partial_t C^{(t)}_{\bs k} +\text{\rm div}(\bs \eta^{(t)}_{\bs k}C^{(t)}_{\bs k}) = 0
\end{align}
for all $t\in [0,t_0]$ where $\bs \eta^{(t)}_{\bs k}:=\bs \eta_{T_t^{-1}(\bs k)}$ and $T_t(\bs k):= \bs k + t\bs \eta_{\bs k}$.
\end{claim}
% --------------------------------------------------------
%  proof
%---------------------------------------------------------
\begin{proof}
By standard optimal transport theory (see \cite{villani2003topics} for example) the assumptions on $C_{\bs k}$ and $\tilde C_{\bs k}$ guarantee the existence of a convex function $\psi_{\bs k}:\Bbb R^d \rightarrow \Bbb R$ such that $\nabla \psi_{\bs k}$ is the optimal transport from $C_{\bs k}d\bs k$ to $\tilde C_{\bs k}d\bs k$.  Let $\bs \eta_{\bs k}:= \frac{1}{t_0}(\nabla \psi_{\bs k}-\bs k)$ so that
\begin{align}
\label{geo eq number 1}
\lambda^{(t)}:=(\bs k + t\bs\eta_{\bs k})\,\sharp\, C_{\bs k} = \big((1-\textstyle\frac{t}{t_0})\bs k + \textstyle\frac{t}{t_0}\nabla\psi_{\bs k}\big)\,\sharp\, C_{\bs k}. 
\end{align}
The particular form of the right hand side of (\ref{geo eq number 1})  implies each measure $\lambda^{(t)}$ has a density $C_{\bs k}^{(t)}$ with respect to Lebesgue measure and the path of measures $\{\lambda^{(t)}\colon t\in[0,t_0] \}$ forms an $L_2$-Wasserstein geodesic with endpoints $C_{\bs k}d\bs k$ and $\tilde C_{\bs k}d\bs k$ at $t=0$ and $t=t_0$ respectively (see Proposition 5.9 in \cite{villani2003topics}). Moreover, $C^{(t)}_{\bs k}$ weakly satisfies (\ref{div flow for geo}) by Theorem 5.34 of \cite{villani2003topics} and the fact that $C^{(t)}_{\bs k} = T_t\,\sharp\, C^{(0)}_{\bs k}$ where $\partial_t T_t(\bs k) = \bs \eta_{\bs k}$.
By  setting $\bs A=t\bs I_d$ in (\ref{eq: local spectral density for nonstationary phase}) one has
\[
\int  \exp\!\big(i(\bs x-\bs y)\!\cdot\! (\bs k + t\bs\eta_{\bs k})\big)\, C_{\bs k}\frac{d\bs k}{(2\pi)^d} 
=\int \exp\!\big(i(\bs x-\bs y)\!\cdot\! \bs\omega\big)\,  C^{(t)}_{\bs \omega}\frac{d\bs \omega}{(2\pi)^d}
\]
which implies that for each $t\in[0,t_0]$ the measure $\lambda^{(t)}$ is a locally attainable spectral measure.
\end{proof}

@@

% --------------------------------------------------------
%  proof
%---------------------------------------------------------
\begin{proof}
By the proof of Claim \ref{claim any two spec are attainable} it will be sufficient to show that $\tilde F^{-1}\circ F(|\bs k|)\frac{\bs k}{|\bs k|}$ is the optimal transport from $C_{|\bs k|}$ to $\tilde C_{|\bs k|}$.
Let $\bs K$ and $\tilde {\bs K}$ be random vectors in $\Bbb R^d$ with densities $C_{|\bs k|}/\sigma^2$ and $\tilde C_{|\bs k|}/\sigma^2$, respectively. By the distributional rotational symmetry of $\bs K$ there exists a convex $\psi(r):\Bbb R^+ \rightarrow \Bbb R$ such that $\nabla (\psi(|\bs k|))=\psi^\prime(|\bs k|)\frac{\bs k}{|\bs k|}$ is the optimal transport from $\mathscr L\bs K$ to $\mathscr L\tilde{\bs K}$ (where $\mathscr L\bs K$ and $\mathscr L\tilde{\bs K}$ denotes the law, i.e. probability distribution, of $\bs K$). Also notice that $\psi^\prime(r)$ is the optimal transport from $\mathscr L |\bs K|$ to  $\mathscr L|\tilde{\bs K}|$ since $\psi(r)$ is convex and
\begin{align*}
P(\psi^\prime(|\bs K|)\leq r) = P\big(\big|\psi^\prime(|\bs K|)\textstyle\frac{\bs K}{|\bs K|}\big|\leq r\big) = P(|\tilde{\bs K}|\leq r).
\end{align*}
The optimal transport between two univariate random variables (see \cite{villani2003topics}) is given by the composition of the quantile function (of the target measure) and the cumulative distribution function (of the base measure). Therefore the optimal transport from $\mathscr L|\bs K|$ to $\mathscr L|\tilde{\bs K}|$ is given by $\tilde F^{-1}\circ F(r)$ where $F(r)=\frac{2\pi^{d/2}}{\sigma^2\Gamma(d/2)}\int_0^r \xi^{d-1}C_\xi d\xi$ and $\tilde F(r)=\frac{2\pi^{d/2}}{\sigma^2\Gamma(d/2)}\int_0^r \xi^{d-1}\tilde C_\xi d\xi$ are the cumulative distribution functions of $|\bs K|$ and $|\tilde{\bs K}|$, respectively. By the uniqueness of optimal transports one has $\psi^\prime(r)=\tilde F^{-1}\circ F(r)$ and therefore 
\[
\nabla (\psi(|\bs k|))=\psi^\prime(|\bs k|)\frac{\bs k}{|\bs k|} = \tilde F^{-1}\circ F(|\bs k|)\frac{\bs k}{|\bs k|}
\]
is the optimal transport from $\mathcal L\bs K$ to $\mathcal L\tilde{\bs K}$,
as was to be shown.
\end{proof}

@@

In this section we give some informal guidance for specifying $\bs \xi_{\bs k}, C_{\bs k}$ and $\bs\eta_{\bs k}$ in the nonstationary spectral phase model (see Definition \ref{def: nonstate phase}). Since the quadratic estimate is  adept at detecting small departures from stationarity---partly due to the accuracy of the variance calculations and the speed at which the quadratic estimate can be simulated under a null stationary model---we focus on the situation where the statistician wants to estimate potential nonstationary extensions of a base stationary model. Further details of this approach can be found in the simulation sections \ref{section: nonstat example d=1} and \ref{section: nonstat example d=2}.
